---
title: "W6_functions"
author: "Aske Svane Qvist"
date: "25 October 2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load packages and data
```{r}
library(pacman)
p_load(tidyverse, gapminder)

# gapminder as df
d <- gapminder

```


# 1. Define a defensive function that calculates the Gross Domestic Product of a nation from the data available in the gapminder dataset. Using that function, calculate the GDP of Denmark in the following years: 1967, 1977, 1987, 1997, 2007, and 2017.

```{r}
# Create function
calcGDP <- function(data, country = NULL, year = NULL){
  
  
  if (!is.na(country)){
    data <- data[data$country %in% country, ]
    # if 'country' is not NA, subset all rows with that or those countries
  }
  
  if (!is.na(year)){
    data <- data[data$year %in% year, ]
  }
  
  gdp <- data$gdpPercap * data$pop
  
  new = cbind(data, gdp)
  
  return(new)
  
}

# Create empty df and make a list of years
DK <- d[0,]
years <- c(1967, 1977, 1987, 1997, 2007)

# loop thorugh all years and gather in a single df
for (i in 1:length(years)){

  new <- calcGDP(d, "Denmark", years[i])
  
  if (length(DK) == 0){
    DK <- new
  } else{
    DK <- rbind(DK, new)
  }

}

# The results
DK

```


# 2. Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a ‘B’ , and print out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r}

# one column df with countries on the list in 2007 (newest)
countries <- d %>% filter(year == 2007)

list <- as.list(countries[,1])
list <- list[["country"]]
length(list)

# Loop through all countries (existing in 2007) and print information
for (i in 1:length(list)){
  
  x <- list[i]
  first <- substr(x, start = 1, stop = 1)
  
  if (isTRUE(first == "B") == TRUE){
    sentence1 <- paste("The first letter of the country", x, "is B.")
  } else {
    sentence1 <- paste("The first letter of the country", x, "is not B.")
  }

  if (countries$lifeExp[i] <= 50){
    sentence2 <- paste("In", x, "the life expectancy was under 50 years in 2007.")
  
  } else if (countries$lifeExp[i] >= 70){
    sentence2 <- paste("In", x, "the life expectancy was greater than 70 years in 2007.")
  
  } else {
    sentence2 <- paste("In", x, "the life expectancy was between 50 and 70 years in 2007.")
  
  }

  print(paste(sentence1,sentence2))
  
}


```

